%!TEX program = pdflatex
%----------------------------------------------------------------------
% Technical Note — The Thermodynamic Cost of Knowing
% Style: UMST thesis-spine voice (monochrome, mathpazo, left-border)
% Compile: cd Docs && pdflatex OnePager-DoubleSlit.tex
%----------------------------------------------------------------------
\documentclass[10pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{mathpazo}
\usepackage[margin=2.0cm, top=1.7cm, bottom=1.8cm]{geometry}
\usepackage{microtype}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{xcolor}
\usepackage{titlesec}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{needspace}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{arrows.meta, positioning}

% --- Monochrome palette (thesis spine) ---
\definecolor{Black}{HTML}{111111}
\definecolor{DimGray}{HTML}{444444}
\definecolor{MidGray}{HTML}{666666}
\definecolor{BorderGray}{HTML}{BBBBBB}
\definecolor{LinkGray}{HTML}{333333}

\renewcommand{\familydefault}{\rmdefault}

\titleformat{\section}
    {\fontsize{11.5}{14}\selectfont\bfseries\color{Black}}
    {\thesection}{0.4em}{}
\titlespacing*{\section}{0pt}{0.65em}{0.2em}

\hypersetup{
    colorlinks=true,
    linkcolor=Black,
    urlcolor=LinkGray,
    citecolor=DimGray,
    pdftitle={The Thermodynamic Cost of Knowing},
    pdfauthor={Santhosh Shyamsundar, Santosh Prabhu Shenbagamoorthy — Studio TYTO}
}

\setlength{\parskip}{0.28em}
\setlist[enumerate]{leftmargin=1.4em, itemsep=0.12em, topsep=0.08em}
\setlist[itemize]{leftmargin=1.4em, itemsep=0.08em, topsep=0.05em}

% Left-border accent box (thesis spine style)
\newenvironment{leftbox}{%
  \par\vspace{0.3em}%
  \noindent\hspace{0pt}%
  \begin{minipage}[t]{3pt}\color{Black}\rule{3pt}{5.5\baselineskip}\end{minipage}%
  \hspace{8pt}%
  \begin{minipage}[t]{\dimexpr\textwidth-15pt}\small
}{%
  \end{minipage}\par\vspace{0.3em}%
}

\newtheorem{theorem}{Theorem}

% Page numbers (bottom center, small, gray)
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\renewcommand{\headrulewidth}{0pt}
\fancyfoot[C]{\footnotesize\color{MidGray}\thepage}

\begin{document}

%---------- HEADER ----------
\noindent
{\fontsize{17}{21}\selectfont\bfseries The Thermodynamic Cost of Knowing}

\vspace{0.15em}
\noindent{\fontsize{12}{15}\selectfont\color{DimGray}\itshape Observation as Irreversible Payment}

\vspace{0.15em}
\noindent{\fontsize{9.5}{12}\selectfont\color{MidGray}%
\textit{Every fraction of a bit of which-path information causes a proportional destruction of interference---and each fraction carries an exact Landauer cost.}}

\vspace{0.25em}
\noindent{\small\color{MidGray}%
A Machine-Checked Proof Connecting Wave-Particle Duality,
Landauer's Erasure Bound, and the Principle of Maximal Information Collapse}

\vspace{0.4em}
\noindent{\small\color{MidGray}\textsf{TECHNICAL NOTE}\quad\rule[0.35ex]{2em}{0.5pt}\quad March 2026}

\vspace{0.2em}
\noindent{\small
\textbf{Santhosh Shyamsundar}\;\;
\textbf{Santosh Prabhu Shenbagamoorthy}}\\[2pt]
{\footnotesize\color{MidGray}%
Studio TYTO (\texttt{tyto.studio})\quad$\cdot$\quad
\texttt{santhosh@tyto.studio}\quad$\cdot$\quad
\textit{Lean\,4 + Mathlib, Haskell QuickCheck, Python}}

\vspace{0.3em}
{\color{Black}\rule{\textwidth}{2.5pt}}

%---------- CORE RESULT ----------
\vspace{0.25em}

\begin{leftbox}
\textbf{Core Result: Principle of Maximal Information Collapse.}
When an observer extracts which-path information from a quantum system, the residual coherence capacity is:
\[
\text{Residual Coherence} = 1 - \frac{\text{MI}_{\text{extracted}}}{k_B T \ln 2} \;\in\; [0,\,1]
\]
Extract 0 bits $\implies$ full interference.\quad Extract 1 bit $\implies$ complete decoherence.\\[2pt]
\textbf{Crucially, observation is not binary.} Partial measurement extracts a \emph{fraction} of a bit, causing proportional decoherence along the Englert curve $V = \sqrt{1 - I^2}$.\\[2pt]
\textit{Machine-checked in Lean\,4 with Mathlib.\; \textbf{360 theorems, 0 sorry, 1 physical axiom.}}
\end{leftbox}

%---------- MOTIVATION ----------
\section{Motivation and Prior Work}

The double-slit experiment is the canonical demonstration of quantum mechanics~\cite{feynman1965, vonneumann1932}: a single particle interferes with itself, producing fringes on a detection screen---unless an observer determines \emph{which slit} the particle traversed, in which case the fringes vanish~\cite{bohr1928}. This complementarity between path information and interference visibility was first quantified by Wootters and Zurek~\cite{wootters1979} and Greenberger and Yasin~\cite{greenberger1988}, then sharpened to Englert's tight inequality $V^2 + I^2 \le 1$~\cite{englert1996}. Though well established, this tradeoff is typically treated as a purely quantum phenomenon, disconnected from classical thermodynamics. The decoherence programme~\cite{zurek2003, schlosshauer2005} explains the \emph{mechanism} of this transition but does not quantify its \emph{thermodynamic cost}.

Our prior work on the \textbf{Unified Material-State Tensor (UMST)} framework~\cite{umst-formal, fcpI, fcpII} established a formally verified thermodynamic gate for continuum materials: a Kleisli-categorical structure enforcing conservation laws (mass, energy, hydration bounds) with graded admissibility. The key insight was that \emph{admissibility is compositional but not transitive}---two individually valid transitions can compose to an inadmissible one, proved via triangle-inequality grading (the \texttt{admissibleTrans} counterexample: densities $\{0, 99, 198\}$ kg/m$^3$ with $\delta = 100$).

The present work extends this framework to quantum measurement. We observe that the act of measurement is not a mysterious ``collapse'' but a \emph{thermodynamic transaction}: extracting which-path information requires physical work bounded by Landauer's principle~\cite{landauer1961, bennett1982} ($Q \ge k_B T \ln 2$ per bit, experimentally confirmed~\cite{berut2012}), and this energy expenditure is \emph{mathematically identical} to the destruction of quantum coherence.

%---------- PROOF ARCHITECTURE ----------
\section{Proof Architecture}

The formalization is structured as a \emph{closed chain} from quantum states through measurement to thermodynamic cost:

\vspace{-0.2em}
\begin{center}
\begin{tikzpicture}[
  box/.style={draw=Black, rounded corners=2pt, minimum width=2.0cm, minimum height=0.6cm, align=center, font=\footnotesize},
  arrow/.style={-{Stealth[length=4pt]}, thick, color=Black}
]
\node[box] (rho) {$\rho \in \mathbb{C}^{2\times 2}$\\PSD, Tr$=1$};
\node[box, right=1.2cm of rho] (iv) {$I = |p_0{-}p_1|$\\$V = 2|\rho_{01}|$};
\node[box, right=1.2cm of iv] (comp) {$V^2 + I^2 \le 1$\\Englert};
\node[box, below=0.8cm of rho] (kraus) {Kraus Channel\\L\"uders $\Pi_i \rho \Pi_i$};
\node[box, below=0.8cm of iv] (ent) {$H = -\sum p_i \ln p_i$\\$\le \ln 2$};
\node[box, below=0.8cm of comp] (land) {$Q \ge k_B T \cdot H$\\Landauer};
\node[box, below=0.8cm of ent, minimum width=3.2cm, line width=0.8pt] (pmic) {\textbf{Residual} $= 1 - S_{\text{bits}}$\;\;$\in [0,1]$};

\draw[arrow] (rho) -- (iv);
\draw[arrow] (iv) -- (comp);
\draw[arrow] (rho) -- (kraus);
\draw[arrow] (kraus) -- node[above, font=\tiny, color=DimGray]{$V\!\to\!0$} (ent);
\draw[arrow] (iv) -- (ent);
\draw[arrow] (ent) -- (land);
\draw[arrow] (land) -- (pmic);
\draw[-{Stealth[length=4pt]}, dashed, thick, color=MidGray] (comp.south west) -- node[below, font=\tiny\itshape, color=MidGray, pos=0.45]{destroys $V$} (kraus.north east);
\end{tikzpicture}
\end{center}
\vspace{-0.3em}

\noindent From the density matrix $\rho$ we extract \textbf{path distinguishability} $I = |p_0 - p_1|$ and \textbf{fringe visibility} $V = 2|\rho_{01}|$. Englert's complementarity follows from the PSD principal minor: $\det(\rho) \ge 0 \implies |\rho_{01}|^2 \le p_0 p_1$, yielding $V^2 + I^2 \le (p_0 + p_1)^2 = 1$.

The L\"uders channel $\mathcal{E}(\rho) = \Pi_0 \rho \Pi_0 + \Pi_1 \rho \Pi_1$ (formalized as Kraus trace-preserving maps with proved self-adjoint, idempotent, orthogonal projectors) zeros the off-diagonals: $V \to 0$, while preserving the diagonal Born weights.

%---------- PMIC ----------
\section{The Principle of Maximal Information Collapse}

The diagonal von Neumann entropy $H = -p_0 \ln p_0 - p_1 \ln p_1 \le \ln 2$ quantifies logical uncertainty. The \textbf{path entropy in bit-equivalents} $S_{\text{bits}} = H / \ln 2 \in [0, 1]$ normalizes this against the single-bit capacity. The \textbf{Landauer cost} is $Q = k_B T \ln 2 \cdot S_{\text{bits}}$, bounded above by one Landauer bit-energy. The \textbf{residual coherence capacity} after extraction:
\[
\text{Residual} = 1 - S_{\text{bits}}
\]
is formally proved to lie in $[0, 1]$, with tight bounds at both endpoints:
\begin{itemize}
\item $S_{\text{bits}} = 0$ (no information extracted) $\implies$ Residual $= 1$ (full visibility possible),
\item $S_{\text{bits}} = 1$ (maximum 1-bit extraction) $\implies$ Residual $= 0$ (complete decoherence).
\end{itemize}
We also prove the cost identity $Q = k_B T \ln 2 \cdot (1 - \text{Residual})$: the thermodynamic cost is \emph{exactly proportional} to the fraction of coherence destroyed. A concrete \texttt{ErasureProcess} structure enforces $\text{dissipatedHeat} \ge \text{landauerCostDiagonal}$ via the Second Law.

\textbf{The collapse is not binary.} The standard textbook narrative presents measurement as all-or-nothing: ``observe and the fringes vanish.'' The Englert relation reveals a \emph{continuum}: a probe that extracts $I = 0.3$ bits of which-path information reduces fringe visibility to $V = \sqrt{1 - 0.09} \approx 0.95$---the interference pattern is barely disturbed. At $I = 0.7$ the fringes are heavily suppressed ($V \approx 0.71$). Full decoherence requires extracting the \emph{entire} bit. Every point on the curve $V^2 + I^2 = 1$ is physically realizable, and each carries a proportional Landauer cost. This continuous tradeoff is precisely what the residual-capacity framework captures and what the Lean formalization proves.

This closes the self-consistent loop: the macroscopic energy expenditure of the intelligent probe maps \emph{directly} to the microscopic destruction of interference.

%---------- THEOREM TABLE ----------
\needspace{10\baselineskip}
\section{What Is Formally Proved}

\noindent Every row is a machine-checked theorem in Lean\,4 with Mathlib~\cite{lean4, mathlib2020}---no \texttt{sorry}, no \texttt{admit}.

\vspace{0.15em}
{\footnotesize
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Theorem} & \textbf{Statement} & \textbf{Module} \\
\midrule
Englert complementarity & $V^2 + I^2 \le 1$ & \texttt{QuantumClassicalBridge} \\
Which-path collapse & $V \to 0$ after L\"uders channel & \texttt{MeasurementChannel} \\
Projector properties & self-adjoint, idempotent, orthogonal, TP & \texttt{MeasurementChannel} \\
Density matrix diagonals & PSD $\implies$ $p_i \ge 0$, $\sum p_i = 1$, $p_i \le 1$ & \texttt{DensityState} \\
Diagonal entropy bound & $H_{\text{diag}} \le \ln 2$ & \texttt{InfoEntropy} \\
Landauer cost cap & cost $\le k_B T \ln 2$ & \texttt{LandauerBound} \\
Path entropy $\le 1$ bit & $S_{\text{bits}} \in [0, 1]$ & \texttt{LandauerBound} \\
Maximal collapse & $S_{\text{bits}} = 1 \implies \text{Residual} = 0$ & \texttt{LandauerBound} \\
Null preservation & $S_{\text{bits}} = 0 \implies \text{Residual} = 1$ & \texttt{LandauerBound} \\
Cost--coherence identity & $Q = k_B T \ln 2 \cdot (1 - \text{Residual})$ & \texttt{LandauerBound} \\
Erasure $\ge$ bound & dissipatedHeat $\ge$ landauerCostDiagonal & \texttt{LandauerBound} \\
Which-path invariance & Landauer cost unchanged by measurement & \texttt{LandauerBound} \\
Gate enforcement & admissibility + Landauer + cap in one & \texttt{DoubleSlit} \\
\bottomrule
\end{tabular}}

%---------- CONNECTION ----------
\section{Connection to the UMST Framework}

This work extends the UMST formal verification stack in three directions:

\begin{enumerate}
\item \textbf{From classical to quantum admissibility.} The original UMST gate checks mass conservation, Clausius-Duhem entropy, hydration bounds on \emph{classical} states (over $\mathbb{Q}$ with \texttt{AdmissibleN}). Here, the gate operates on \emph{quantum} density matrices: the projector $\Pi_i \rho \Pi_i$ is the measurement-channel analogue, and admissibility is enforced by Englert complementarity plus the Landauer bound.

\item \textbf{From material hydration to information erasure.} In the original framework, the Helmholtz free energy $\psi(\alpha)$ governs the cost of hydration transitions. Here, the diagonal von Neumann entropy $H(\rho)$ governs the cost of information extraction. Both are monotone hooks: increasing hydration lowers free energy; increasing path information lowers residual coherence. The Galois adjunction \texttt{EpistemicGalois} makes this connection explicit.

\item \textbf{From counterexample to principle.} The discovery that \texttt{admissibleTrans} is \emph{refutable} led to graded admissibility. The analogous quantum insight is that \emph{partial} measurements compose non-trivially: two weak measurements each extracting $< 1$ bit can together extract the full bit, collapsing coherence entirely. The residual-capacity framework captures this compositional structure.
\end{enumerate}

%---------- VERIFICATION ----------
\section{Verification Stack}

{\footnotesize
\begin{tabular}{@{}llll@{}}
\toprule
\textbf{Layer} & \textbf{Tool} & \textbf{Scope} & \textbf{Status} \\
\midrule
Formal proof & Lean 4 + Mathlib & 38 modules, 360 theorems & \textbf{0 sorry} \\
Property testing & Haskell QuickCheck & 14 properties + sanity & \textbf{All pass} \\
Numerical sim & Python (stdlib + QuTiP) & 54 unit tests, 4 sims & \textbf{All pass} \\
Cross-language & Coq + Agda & LandauerEinsteinBridge & \textbf{0 Admitted} \\
\bottomrule
\end{tabular}}

\vspace{0.15em}
\noindent The Python simulator demonstrates the collapse visually: as which-path information $I$ increases from 0 to 1, the interference pattern progressively flattens, tracking the $V = \sqrt{1 - I^2}$ Englert curve exactly.

\vspace{0.25em}
{\color{Black}\rule{\textwidth}{1pt}}
\vspace{0.1em}

\noindent{\small\textit{Repository:} \texttt{umst-formal-double-slit} \quad\textbar\quad \textit{License:} MIT \quad\textbar\quad \textit{Build:} \texttt{cd Lean \&\& lake build}}

\noindent{\footnotesize\color{MidGray}\copyright\;2026 Studio TYTO $\cdot$ \texttt{github.com/tytolabs}}

\vspace{0.5em}
\noindent{\footnotesize\color{MidGray}\textit{Acknowledgments.}\;
Portions of this work were developed with assistance from large-language-model tools
(Claude by Anthropic, Gemini by Google, Grok by xAI) and the Cursor code editor.
All formal proofs were machine-checked by their respective compilers (Lean\,4, Coq, Agda);
the LLMs contributed to exploration, drafting, and code scaffolding---not to proof validity.}

%---------- REFERENCES ----------
\begingroup
\small
\begin{thebibliography}{99}

%--- Our work ---
\bibitem{umst-formal} S.~Shyamsundar and S.\,P.~Shenbagamoorthy, ``Formal Verification and Categorical Semantics for the UMST/DUMSTO Thermodynamic Gate,'' \texttt{umst-formal}, GitHub, 2026. 59 Lean theorems, 0 sorry.

\bibitem{fcpI} S.~Shyamsundar and S.\,P.~Shenbagamoorthy, ``Towards Unified Material-State Tensors for Physics-Gated AI: Thermodynamic Admissibility as Constitutional Constraint,'' Zenodo, 2026. \texttt{doi:10.5281/zenodo.18768547}

\bibitem{fcpII} S.\,P.~Shenbagamoorthy and S.~Shyamsundar, ``Epistemic Sensing Architecture for Physics-Constrained Material Characterization,'' Zenodo, 2026. \texttt{doi:10.5281/zenodo.18894710}

%--- Quantum foundations ---
\bibitem{bohr1928} N.~Bohr, ``The quantum postulate and the recent development of atomic theory,'' \textit{Nature}, 121:580--590, 1928.

\bibitem{feynman1965} R.\,P.~Feynman, R.\,B.~Leighton, and M.~Sands, \textit{The Feynman Lectures on Physics}, Vol.\,III, Ch.\,1, Addison-Wesley, 1965.

\bibitem{vonneumann1932} J.~von~Neumann, \textit{Mathematische Grundlagen der Quantenmechanik}, Springer, 1932.

\bibitem{wootters1979} W.\,K.~Wootters and W.\,H.~Zurek, ``Complementarity in the double-slit experiment,'' \textit{Phys.\ Rev.\ D}, 19(2):473--484, 1979.

\bibitem{greenberger1988} D.\,M.~Greenberger and A.~Yasin, ``Simultaneous wave and particle knowledge in a neutron interferometer,'' \textit{Phys.\ Lett.\ A}, 128:391--394, 1988.

\bibitem{englert1996} B.-G.~Englert, ``Fringe visibility and which-way information: An inequality,'' \textit{Phys.\ Rev.\ Lett.}, 77(11):2154--2157, 1996.

%--- Thermodynamics of information ---
\bibitem{landauer1961} R.~Landauer, ``Irreversibility and heat generation in the computing process,'' \textit{IBM J.\ Res.\ Dev.}, 5(3):183--191, 1961.

\bibitem{bennett1982} C.\,H.~Bennett, ``The thermodynamics of computation---a review,'' \textit{Int.\ J.\ Theor.\ Phys.}, 21:905--940, 1982.

\bibitem{berut2012} A.~B\'{e}rut \textit{et al.}, ``Experimental verification of Landauer's principle linking information and thermodynamics,'' \textit{Nature}, 483:187--189, 2012.

%--- Decoherence ---
\bibitem{zurek2003} W.\,H.~Zurek, ``Decoherence, einselection, and the quantum origins of the classical,'' \textit{Rev.\ Mod.\ Phys.}, 75:715--775, 2003.

\bibitem{schlosshauer2005} M.~Schlosshauer, ``Decoherence, the measurement problem, and interpretations of quantum mechanics,'' \textit{Rev.\ Mod.\ Phys.}, 76:1267--1305, 2005.

\bibitem{nielsen2000} M.\,A.~Nielsen and I.\,L.~Chuang, \textit{Quantum Computation and Quantum Information}, Cambridge Univ.\ Press, 2000.

%--- Formal verification ---
\bibitem{lean4} L.~de~Moura and S.~Ullrich, ``The Lean~4 Theorem Prover and Programming Language,'' \textit{CADE-28}, pp.\,625--635, Springer, 2021.

\bibitem{mathlib2020} The Mathlib Community, ``The Lean Mathematical Library,'' \textit{CPP 2020}, pp.\,367--381, ACM, 2020.

\bibitem{avigad2017} J.~Avigad, J.~H\"{o}lzl, and L.~Serafin, ``A formally verified proof of the central limit theorem,'' \textit{J.\ Automated Reasoning}, 59:389--423, 2017.

\end{thebibliography}
\endgroup

\end{document}
